A new method to explicitly estimate the shift of optimum along gradients in multispecies studies

### 2022-11-04 redaction/manuscript_cut.Rmd
Corrections after revisions by JoB reviewers

---
title: "A new method to explicitly estimate the shift of optimum along gradients in multispecies studies"
subtitle: "Running title: Modelling of gradient optimum shifts"
classoption: a4paper
output:
  pdf_document:
    keep_tex: true
  word_document: default
  documentclass: article
fontsize: 12pt
header-includes:
- \usepackage{float}
- \usepackage{lscape}
- \usepackage{amsmath}
- \usepackage{caption}
- \usepackage{lineno}
- \linenumbers
#- \usepackage{setspace}\doublespacing
csl: journal-of-biogeography.csl
bibliography: Paper_EHMOS_cut.bib
---

@@

## Abstract
### Aim
Optimum shifts in species–environment relationships are intensively studied in a wide range of ecological topics, including climate change and species invasion. Numerous statistical methods are used to study optimum shifts but, to our knowledge, none explicitly estimate it. We extended an existing model to explicitly estimate optimum shifts for multiple species. We called this new Bayesian hierarchical model the Explicit Hierarchical Model of Optimum Shifts (EHMOS).

@@

Understanding changes in relationships between species and environmental, geographical or temporal gradients is of central interest in biogeography. For instance, assessing changes in species distribution along latitudinal [@thorson2016], elevational [@maggini2011] or climatic [@tayleur2015] gradients in response to ongoing climate change is crucial for developing relevant biodiversity change scenarios [@taheri2021] and predicting its impact on ecosystem health and human well-being [@pecl2017]. Modelling these changes is also central in the study of invasive species to predict the area of potential colonization [@elith2010]. The shifts in species phenology, i.e. changes along a temporal gradient, in response to climate change are also well-studied to obtain a mechanistic understanding of climate change impacts on species distributions [@moussus2010; @vitasse2021; @strebel2014].

All this research aims at modelling the relationship between species characteristics –  such as occurrence probability [@maggini2011], abundance [@thorson2016] or singing activity [@strebel2014] – and linear covariates describing the studied gradient. These modelled relationships, often named "species response curves" [@tayleur2015], are usually unimodal curves [@oksanen2002]. They are characterized by a width, a maximum, an optimum (i.e. the value on the gradient where the maximum abundance or occurrence probability of the species is reached), and a lower and upper limit.  Changes in the relationship could then occur in each of these parameters, providing different and complementary information on the way species are affected by change over the gradient. However, these changes may be difficult to estimate. Changes in width or range limits of the species response curve may for instance be more difficult to estimate than optimum shifts due to sampling design or detectability issues linked to lower abundance near range limits [@shoo2006]. Hence, studying optimum shifts is often favoured over shifts at range limits to disentangle ecological changes from potential sampling artifacts [@shoo2006; @moussus2010; @lenoir2008].

Different modelling frameworks have been developed to study optimum shifts. The simplest, and probably the most widely used, is the mean comparison method [e.g. @shoo2006; @freeman2018], also known as the weighted average method [@terbraak1986; @thorson2016], in which mean values of occupied sites on the gradient are compared between two or more samples. However, such optimum estimates could be biased when sampling effort is uneven along the gradient [@shoo2006; @terbraak1986]. Such uneven sampling is common in ecology [@shoo2006; @rumpf2018], especially in cases of sampling difficult-to-access terrains such as mountains. Designing a random sample of a species distribution requires knowing the species distribution prior to sampling it, which is almost never the case. Regression methods, such as generalized linear models (GLMs), are also commonly used to estimate optimum shifts [@urli2014; @coudun2005; @lenoir2008] and are known to be more robust to unbalanced sampling design [@terbraak1986]. In this approach, two regressions are conducted to estimate the species optimum for each sample separately [@coudun2005]. Shift is then calculated as the difference between the two estimated optima [as in @lenoir2008]. Such a separate analysis makes it difficult to estimate uncertainty around the shift, so this uncertainty is often omitted [e.g. @lembrechts2017; @urli2014]. This absence of uncertainty is particularly prejudicial when post-hoc tests are performed on shift point estimates [@felde2012]. This approach also raises issues for species with an optimum close to the edge of the sampled gradient, hereafter referred to as edge species, also called marginal species [@hernandez2006]. In these cases, the regression follows a logistic relationship and does not provide an optimum [@citores2020]. As a consequence, such edge species are usually withdrawn from analysis [e.g. @rumpf2018; @felde2012]; however, they may be particularly sensitive to changes as they are limited by certain constraints (physiological, environmental, etc.) and subject to extirpation phenomenon, i.e. local extinctions caused by a lack of remaining suitable areas [@freeman2018]. Moreover, estimates of the mean shift at the community level could be biased by the removal of edge species, as the optimum shift could depend on the position along the gradient [@rumpf2018]. A model robust to unbalanced sampling design, providing estimates of uncertainties and including all species whatever the location of their optimum would be a substantial improvement for modelling shift in optima.

We developed an extension of the Gaussian logistic model [@jamil2014; @terbraak1986] that explicitly estimates shifts in optima for several species simultaneously, with their uncertainties. We also fitted a Bayesian Generalized Linear Mixed Model to simultaneously analyse both sampling occasions and all species and use the posterior distributions of the estimated coefficients to compute posterior distributions of species optimum shifts. The performance of these models was evaluated and compared to the estimates provided by a simple mean comparison method. The comparison was undertaken through a simulation study involving different sampling designs and ecological scenarios. The primary goal of the simulation study was to determine if the tested methods accurately estimate the optimum shifts of multiple species along a virtual environmental gradient. Specifically, we tested if the accuracy of the methods was sensitive to (1) sampling design, i.e. how sampling sites are distributed along the gradient, (2) species marginality, i.e. the location of the species optimum on the gradient, and (3) species ecological specialization, i.e. the width of the species–environment relationship. We finally applied the three methods to 24 Orthoptera species sampled twice 30 years apart in the French Alps along an elevation gradient.

@@

We used a mean comparison method based on the Student t-test (hereafter, t-test) that compares species' average positions along a gradient between two sampling occasions, e.g. two periods or two regions. In this approach, species' optima are estimated as the mean of environmental values at which the species have been observed. We performed unpaired two-sample Student t-tests between surveys for each species separately. This provided species-specific shift estimates and associated confidence intervals.

@@

We assumed that species occupancy probability is related to the gradient through a symmetric bell-shaped curve, with an optimum where the maximum occupancy probability is reached. However, the two models differ in the formulation of the species–environment relationship. cGLMM allows five shapes of species-environment relationships (flat, monotone, 'plateau', U-shaped or bell-shaped unimodal symmetric response curves) some of which could be ecologically meaningless (e.g. U-shaped curves). EHMOS by its ecological formulation models only symmetric unimodal response curves. However, note that if EHMOS estimates optimum outside the sampling range, then the estimated SRC will look like a monotone or a 'plateau' relationship. EHMOS ecological formulation also allows direct interpretation of its model coefficients, especially about optimum and shift values. Note that both models assumed that unimodal species response curves are symmetrical, i.e. occupancy probabilities decrease at the same rate on both sides of the optimum. We discuss this potential limitation further in the manuscript.

@@

Our goal when starting this study was to find an appropriate modelling approach that produces estimates, and associated uncertainties, of shifts in elevation optima of an Orthoptera community studied during two periods separated by 30 years. This study was conducted in 134 sampling sites distributed along an elevation gradient ranging from 928 m to 2614 m (mean = 1869 m) above sea level in Mercantour National Park in the southern French Alps. The first survey was carried out during the summers between 1983 and 1988 (Gueguen 1990). At each site, the density of Orthoptera species was assessed based on multiple counts made using box quadrats that trapped individuals. The second survey was conducted during the summers of 2018 and 2019. Each sampling site consisted of a circle with a 70-m radius where five spatial replicates (hereafter, plots) 30 m2 in area were surveyed following three successive steps: (1) one minute of listening to species stridulating in the plot by standing close to its edge, (2) six minutes of sighting species by walking across the entire plot, and (3) two 45-second sweep netting sessions across the entire plot [see @mourguiart2021 for a full description of the sampling design]. To minimize the effect of varying sampling effort and detection methods between surveys, we reduced the Orthoptera records to the presence/absence data at the site level and only kept species detected in at least 5\% of sampling sites. We pooled some species' data in a species complex when species confusion was possible. Thus, we aggregated data for species in the genuses *Yersinella*, *Calliptamus*, *Podisma* and in the complex *Chorthippus biggutulus-brunneus-mollis-daimei*. We fitted the three methods to the 24 taxa kept after data processing.

@@

Before analysing the results, we checked for convergence issues and running times of cGLMM and EHMOS. In the $\hat{R}$ statistics check, we observed that 50\% of runs failed to converge (at least one parameter had $\hat{R}$ higher than 1.1) for cGLMM (see in Appendix S3 as Table S3.1). Due to the running time, `r m.compute_time[m.compute_time$model == "glmm", ]$mean` hours (SD = `r m.compute_time[m.compute_time$model == "glmm", ]$SD` hours) on average for cGLMM (see in Appendix S3 as Table S3.2), we chose not to re-run more iterations of models that did not converge. All models converged for EHMOS and took on average `r m.compute_time[m.compute_time$model == "ehmos", ]$mean` hours (SD = `r m.compute_time[m.compute_time$model == "ehmos", ]$SD` hour) for this process (see in Appendix S3 as Table S3.2). In the motivating example, both models converged: EHMOS ran in `r round(out_appli_ehmos$mcmc.info$elapsed.mins, 0)` minutes and cGLMM in `r round(out_appli_glmm$mcmc.info$elapsed.mins, 0)` minutes.

@@

The three methods provided heterogeneous shift estimates (Fig. \ref{fig:fig3}). T-test estimates seemed to be smaller than those estimated by cGLMM or EHMOS. To verify this observation, we performed a paired t-test on species point estimates between methods. We found significant differences between the t-test and cGLMM estimates (mean difference = `r round(appli_ttest.vs.glmm$estimate, 0)`, t(`r appli_ttest.vs.glmm$parameter`) = `r round(appli_ttest.vs.glmm$statistic, 2)`, p = `r round(appli_ttest.vs.glmm$p.value, 2)`) and EHMOS estimates (mean difference = `r round(appli_ttest.vs.ehmos$estimate, 0)`, t(`r appli_ttest.vs.ehmos$parameter`) = `r round(appli_ttest.vs.ehmos$statistic, 2)`, p < 0.01), confirming our observation on the graph. Methods also differed in the precision of estimates. The t-test had on average narrower confidence intervals with a mean confidence interval width of `r appli_mean_CI.shift[appli_mean_CI.shift$model == "ttest", ]$m.CI.shift %>% round(0) ` m, compared to EHMOS (`r appli_mean_CI.shift[appli_mean_CI.shift$model == "ehmos", ]$m.CI.shift %>% round(0) ` m) and cGLMM (`r appli_mean_CI.shift[appli_mean_CI.shift$model == "glmm", ]$m.CI.shift %>% round(0) ` m). This high value for cGLMM is due to unreliably wide credible intervals for nine species (Fig. \ref{fig:fig3}). Those unrealistic estimates could be explained by quadratic coefficients, the denominator in optimum formulae (Eq. \ref{shift_glmm}), estimated close to zero and generating unreliable optimum estimates and thus shift estimates. This seemed to occur for taxa having their optimum close to the sampling edge (before the 10^th^ or beyond the 90^th^ percentile of the sampling site elevations, depicted as orange rectangles in Fig. \ref{fig:fig3}), for example,  *Yersinella sp.* and *Gomphocerus sibiricus sibiricus*, or species having low maximum occupancy probability, e.g. *Omocestus haemorrhoidalis*.

@@

EHMOS always performed better than the t-test-based method. Differences between these two methods were particularly large under unbalanced sampling design, and especially in the presence of generalist species. It is predictable that mean estimates are biased by unbalanced sampling designs, in particular in scenarios where a part of the gradient is preferentially sampled leading to bias in optimum estimates towards this part [@terbraak1986]. We simulated one example where the middle part of the gradient was preferentially sampled, but similar results are expected in other preferential sampling design, e.g. preferential sampling towards low elevations due to the reduction in surface area with elevation [@lenoir2008]. In such situations, both optima estimates could be biased towards the preferentially sampled part of the gradient and shift estimates could be biased towards zero. This explains the negative bias produced by t-test method in our simulated scenarios involving unbalanced sampling design. It is also not surprising that bias due to unbalanced sampling design is larger for generalist species than for specialist species. Bias is more influenced by the distribution of sampling sites occupied by the species rather than by the entire distribution of sampling sites. Occupied sites of a specialist species with a narrow range could be uniformly distributed even if the entire distribution of sampling sites is uneven. In contrast, the larger the species range, the more the occupied site distribution on the gradient will match the entire sampling site distribution, leading to biased shift estimates.

@@

Generalized linear models have been extensively used in species distribution modelling [@elith2010], and also in optimum shift modelling [e.g. @coudun2005; @lenoir2008]. Usually, multiple GLMs are conducted separately for each species and each survey that is to be compared [but see @lembrechts2017 using GLMM]. This  approach has the inconvenient of losing information and thus precision in estimation by splitting the data into multiple data sets. Estimates for data-poor species may for instance be improved with a GLMM, by borrowing information from data-rich species [@ovaskainen2011]. Hence, one could assume that GLM would have, at best equal and probably worse performances than cGLMM.  We thus chose to compare EHMOS only with cGLMM to show that EHMOS has advantages against classical GLM-based approaches. We worked with a Bayesian framework that allows us to derive uncertainty measures associated with shift estimates. Note that it would have been possible to compute the asymptotic variance of optima or shift estimates using a frequentist approach, e.g. the delta method [@urli2014]. The Bayesian GLMM we developed here performed as well as the EHMOS and better than the mean comparison method in scenarios where all species have optima in the middle of the sampling range. However, as expected, for edge species, the standard GLMM failed to estimate bell shape curves, fitting sigmoid curves instead and making derived optima and shifts unreliable [@citores2020; @terbraak1986]. In contrast, EHMOS, thanks to its ecological formulation and its explicit modelling of optimum shifts, allowed estimates of optimum shifts for edge species. Hence, edge species could be kept if one is using EHMOS, while they should be removed in t-test or GLM-based analysis. Ecological formulation of EHMOS, especially explicit modelling of shift, will also allow easier choice of prior in a Bayesian setting. cGLMM parameters should be transformed to be ecologically interpretable (see Appendix S1). Such a transformation may change an uninformative prior on the parameter scale into an informative prior on the ecological scale, making the choice of prior more complicated [see @lemoine2019 for an example in logistic regression]. Hence, the standard GLMM appeared to be less flexible than EHMOS.

In addition to having more robust point estimate accuracy whatever the sampling design, ecological marginality and specialization than a t-test or cGLMM, EHMOS also seemed to produce more precise estimates. cGLMM was very sensitive to ecological marginality in terms of precision, providing very wide and uninformative credible intervals for edge species. The interval coverage of the t-test was also very sensitive to ecological marginality, sampling design and ecological specialization. Having informative precision estimates and reporting them is important when studying species shifts [@bates2015; @taheri2021]. Some authors use the overlap of confidence intervals as a surrogate for a test of significant difference [@lenoir2008; @urli2014], so low interval coverage could lead to misleading conclusions, and the t-test method should be avoided.

In our case study, we studied optimum shifts between two surveys conducted 30 years apart of 24 Orthoptera species along an elevational gradient in the French Alps. All methods suggested that on average Orthoptera species shifted their optimum towards higher elevations, and that the magnitude of the shift varies between species. These results are consistent with previous findings on other insect taxa, which on average have shifted upslope, but at different paces depending on the species [@vitasse2021]. Even if the magnitude of shifts varied between methods, our estimates are in line with the estimated warming rate of 0.36 $^{\circ}$C.decade^-1^ in the Alps between 1970 and 2019, which roughly corresponds to a mean shift of about +62–71 m decade^–1^ (Vitasse et al. 2021), leading to an expected mean shift of around +186–213 m between the two surveys. Thus, the average shift estimated by EHMOS (`r appli_mean_shift[appli_mean_shift$model == "ehmos", ]$m.shift %>% round(0)` m) is closer to what is expected than the t-test, suggesting it may provide a more reliable estimate of optimum shift. Yet this observation has to be taken carefully, as the warming rate could vary between regions, and species might respond at a slower rate than expected [@vitasse2021]. However, as the case study is close to the scenario A2xB2xC2 of our simulations, with the presence of both generalist and edge species sampled with an unbalanced sampling design, we could suspect an underestimation of optimum shifts by the t-test as observed in these simulations. It was also not surprising to obtain unreliable wide credible intervals for most of the edge species with the GLMM, as the simulations provide such evidence.

Notwithstanding its demonstrated comparative advantages, EHMOS could have some limitations. It might produce optimum estimates outside the sampling range and thus, depending on the sampling design, outside the environmental range. This property could be an advantage in the case of niche truncation. However, optimum estimates should be limited to the environmental range: for example, by imposing an appropriate prior. One potential other limitation, also valid for cGLMM, is the assumption that species response curves are symmetric. While this assumption could stand for numerous species [@rydgren2003], some species could have asymmetrical relationships with environment gradients [@oksanen2002]. We found through a simulation study that the three methods could be biased in presence of asymmetrical species response curves (see Appendix S4). More precisely, the three methods provided biased optimum shift estimates when species response curves changed of shapes between the sampling occasions. For instance, the three methods produced positively biased optimum shift estimates for species having a symmetrical response curve in the first sampling occasion and a right-skewed species response curve in the second sampling occasion,  situations that could arise due to range margin contraction or expansion for instance. In contrast, optimum shift should be accurately estimated by the three methods for species having the same degree of skewness in both sampling occasions. Hence, shape of species response curves should be inspected carefully prior to analysis. In case where response curves are asymmetric, some alternative modelling methods have been proposed, such as generalized additive models [GAM, e.g. @heegaard2002] or Huisman-Olff-Fresco approach [HOF, e.g. @oksanen2002], and applied in optimum shift modelling [@maggini2011; @urli2014]. However, previous simulation studies have shown that both methods could fail to retrieve the true shape of species response [@jansen2013; @michaelis2017]. Both methods appeared to be sensitive to gradient truncation and sampling intensity [@jansen2013; @citores2020], limiting their use in unbalanced sampling design and for edge species. Further research is thus needed to find more robust methods to detect and deal with asymmetrical response curves. Finally, we present the EHMOS within a hierarchical model that did not incorporate observation process. It is well known that imperfect detection could bias occupancy estimates, and that this should be addressed in modelling whenever it is possible. While our example does not allow the use of multi-species occupancy models, hierarchical linear models estimating both species occupancy and detection [@dorazio2005], due to the absence of sampling replication in historical data, it is straightforward to extend our framework in a MSOM by adding the observation process in the hierarchy of the model.

>>>>

---
title: "A new method to explicitly estimate the shift of optimum along gradients in multispecies studies"
subtitle: "Running title: Modelling of gradient optimum shifts"
classoption: a4paper
output:
  pdf_document:
    keep_tex: true
  word_document: default
  documentclass: article
fontsize: 12pt
header-includes:
- \usepackage{float}
- \usepackage{lscape}
- \usepackage{amsmath}
- \usepackage{caption}
- \usepackage{lineno}
- \linenumbers
- \usepackage{setspace}\setstretch{1.5}
csl: journal-of-biogeography.csl
bibliography: Paper_EHMOS_cut.bib
---

@@

## Abstract
### Aim
Optimum shifts in species–environment relationships are intensively studied in a wide range of ecological topics, including climate change and species invasion. Numerous statistical methods are used to study optimum shifts but, to our knowledge, none explicitly estimate it. We extended an existing model to explicitly estimate optimum shifts for multiple species having symmetrical response curves. We called this new Bayesian hierarchical model the Explicit Hierarchical Model of Optimum Shifts (EHMOS).

@@

Understanding changes in relationships between species and environmental gradients is of central interest in biogeography. For instance, assessing changes in species distribution along latitudinal [@thorson2016], elevational [@maggini2011] or climatic [@tayleur2015] gradients in response to ongoing climate change is crucial for developing relevant biodiversity change scenarios [@taheri2021]. Modelling these changes is also central in assessing species invasion risks [@elith2010]. The shifts in species phenology, i.e. changes along a temporal gradient, in response to climate change are also well-studied [@moussus2010; @vitasse2021; @strebel2014].

This research aims at modelling the relationship between species characteristics –  such as occurrence probability [@maggini2011], abundance [@thorson2016] or singing activity [@strebel2014] – and linear covariates describing the studied gradient. These modelled relationships, often named "species response curves" [@tayleur2015], are usually unimodal curves [@oksanen2002]. They are characterized by a width, a maximum, an optimum (i.e. the value on the gradient where the response peak is reached), and a lower and upper limit.  Changes in the relationship could then occur in each of these parameters, providing different and complementary information on the way species are affected by change over the gradient. However, these changes may be difficult to estimate. Changes in width or range limits of the species response curve may for instance be more difficult to estimate than optimum shifts due to sampling design or detectability issues linked to lower abundance near range limits [@shoo2006]. Hence, studying optimum shifts is often favoured over shifts at range limits to disentangle ecological changes from potential sampling artifacts [@shoo2006; @moussus2010; @lenoir2008].

Different modelling frameworks have been developed to study optimum shifts. The simplest is the mean comparison method [e.g. @shoo2006; @freeman2018], also known as the weighted average method [@terbraak1986; @thorson2016], in which mean values of occupied sites on the gradient are compared between two or more samples. However, such optimum estimates could be biased when sampling effort is uneven along the gradient [@shoo2006; @terbraak1986]. Such uneven sampling is common in ecology [@shoo2006; @rumpf2018], especially in cases of sampling difficult-to-access terrains such as mountains. Designing a random sample of a species distribution requires knowing the species distribution prior to sampling it, which is seldom the case. Regression methods, such as generalized linear models (GLMs), more robust to unbalanced sampling design [@terbraak1986], are also commonly used to estimate optimum shifts [@urli2014; @coudun2005; @lenoir2008]. In this approach, two regressions are conducted to estimate the species optimum for each sample separately [@coudun2005]. Shift is calculated as the difference between the two estimated optima [as in @lenoir2008]. Such an analysis makes it difficult to estimate uncertainty around the shift, so this uncertainty is often omitted [e.g. @lembrechts2017; @urli2014]. This absence of uncertainty is prejudicial when post-hoc tests are performed on shift point estimates [@felde2012]. This approach also raises issues for species with an optimum close to the edge of the sampled gradient, hereafter referred to as edge species. In these cases, the regression follows a logistic relationship and does not provide an optimum [@citores2020]. As a consequence, such edge species are usually withdrawn from analysis [e.g. @rumpf2018; @felde2012]; however, they may be particularly sensitive to changes as they are limited by certain constraints (physiological, environmental, etc.) and subject to local extinctions caused by a lack of remaining suitable areas [@freeman2018]. Moreover, estimates of the mean shift at the community level could be biased by the removal of edge species, as the optimum shift could depend on the position along the gradient [@rumpf2018]. A model robust to unbalanced sampling design, providing estimates of uncertainties and including all species whatever the location of their optimum would be a substantial improvement for modelling shift in optima.

We developed an extension of the Gaussian logistic model [@jamil2014; @terbraak1986] that explicitly estimates shifts in optima for several species simultaneously, with their uncertainties. We also fitted a Bayesian Generalized Linear Mixed Model to simultaneously analyse both sampling occasions and all species and use the estimated coefficients to compute posterior distributions of species optimum shifts. The performance of these models was compared to the estimates provided by a simple mean comparison method. The comparison was undertaken through a simulation study involving different sampling designs and ecological scenarios. The goal of the simulation study was to determine if the methods accurately estimate the optimum shifts of multiple species along a virtual environmental gradient. Specifically, we tested if the accuracy of the methods was sensitive to (1) sampling design, i.e. how sampling sites are distributed along the gradient, (2) species marginality, i.e. the location of the species optimum on the gradient, and (3) species ecological specialization, i.e. the width of the species–environment relationship. We finally applied the three methods to Orthoptera species sampled twice 30 years apart in the French Alps along an elevation gradient.

@@

We used a mean comparison method based on the Student t-test (hereafter, t-test) that compares species' average positions along a gradient between two sampling occasions. In this approach, species' optima are estimated as the mean of environmental values at which the species have been observed. We performed unpaired two-sample Student t-tests between surveys for each species separately. This provided species-specific shift estimates and associated confidence intervals.

@@

We assumed that species occupancy probability is related to the gradient through a symmetric bell-shaped curve with an optimum. However, the two models differ in the formulation of the species–environment relationship. cGLMM allows five shapes of species-environment relationships (flat, monotone, 'plateau', U-shaped or bell-shaped unimodal symmetric response curves) some of which could be ecologically meaningless (e.g. U-shaped curves). EHMOS by its ecological formulation models only symmetric unimodal response curves. However, if EHMOS estimates optimum outside the sampling range, then the estimated SRC will look like a monotone or a 'plateau' relationship. EHMOS ecological formulation also allows direct interpretation of its model coefficients. Note that both models assumed that unimodal species response curves are symmetrical, i.e. occupancy probabilities decrease at the same rate on both sides of the optimum. We discuss this potential limitation further in the manuscript.

@@

Our goal when starting this study was to model shifts in elevation optima of an Orthoptera community studied during two periods separated by 30 years. This study was conducted in 134 sampling sites distributed along an elevation gradient ranging from 928 m to 2614 m (mean = 1869 m) above sea level in Mercantour National Park in the French Alps. The first survey was carried out during the summers between 1983 and 1988 (Gueguen 1990). At each site, the density of Orthoptera species was assessed based on multiple counts made using box quadrats that trapped individuals. The second survey was conducted during the summers of 2018 and 2019. Each sampling site consisted of a circle with a 70-m radius where five spatial replicates (hereafter, plots) 30 m2 in area were surveyed following three successive steps: (1) one minute of listening to species stridulating in the plot, (2) six minutes of sighting species by walking across the plot, and (3) two 45-second sweep netting sessions across the plot [see @mourguiart2021 for a full description of the sampling design]. To minimize the effect of varying sampling effort and detection methods between surveys, we reduced the Orthoptera records to the presence/absence data at the site level and only kept species detected in at least 5\% of sampling sites. We pooled some species' data in a species complex when confusion was possible (genuses *Yersinella*, *Calliptamus*, *Podisma* and *Chorthippus biggutulus-brunneus-mollis-daimei* complex). We fitted the three methods to the 24 taxa kept after data processing.

@@

In the convergence check, we observed that 50\% of the cGLMMs failed to converge in the simulations (see details in Appendix S3 as Table S3.1). Due to the running time, `r m.compute_time[m.compute_time$model == "glmm", ]$mean` hours (SD = `r m.compute_time[m.compute_time$model == "glmm", ]$SD` hours) on average for cGLMM (see in Appendix S3 as Table S3.2), we chose not to re-run more iterations of models that did not converge. All models converged for EHMOS and took on average `r m.compute_time[m.compute_time$model == "ehmos", ]$mean` hours (SD = `r m.compute_time[m.compute_time$model == "ehmos", ]$SD` hour) for this process (see in Appendix S3 as Table S3.2). In the motivating example, both models converged: EHMOS ran in `r round(out_appli_ehmos$mcmc.info$elapsed.mins, 0)` minutes and cGLMM in `r round(out_appli_glmm$mcmc.info$elapsed.mins, 0)` minutes.

@@

The three methods provided heterogeneous shift estimates (Fig. \ref{fig:fig3}). T-test estimates seemed to be smaller than those estimated by cGLMM or EHMOS. To verify this observation, we performed a paired t-test on species point estimates between methods. We found significant differences between the t-test and cGLMM estimates (mean difference = `r round(appli_ttest.vs.glmm$estimate, 0)`, t(`r appli_ttest.vs.glmm$parameter`) = `r round(appli_ttest.vs.glmm$statistic, 2)`, p = `r round(appli_ttest.vs.glmm$p.value, 2)`) and EHMOS estimates (mean difference = `r round(appli_ttest.vs.ehmos$estimate, 0)`, t(`r appli_ttest.vs.ehmos$parameter`) = `r round(appli_ttest.vs.ehmos$statistic, 2)`, p < 0.01), confirming our visual observation. Methods also differed in the precision of estimates. The t-test had on average narrower confidence intervals with a mean confidence interval width of `r appli_mean_CI.shift[appli_mean_CI.shift$model == "ttest", ]$m.CI.shift %>% round(0) ` m, compared to EHMOS (`r appli_mean_CI.shift[appli_mean_CI.shift$model == "ehmos", ]$m.CI.shift %>% round(0) ` m) and cGLMM (`r appli_mean_CI.shift[appli_mean_CI.shift$model == "glmm", ]$m.CI.shift %>% round(0) ` m). This high value for cGLMM is due to unreliably wide credible intervals for nine species (Fig. \ref{fig:fig3}). Those unrealistic estimates could be explained by quadratic coefficients, the denominator in optimum formulae (Eq. \ref{shift_glmm}), estimated close to zero and generating unreliable optimum estimates and thus shift estimates. This seemed to occur for taxa having their optimum close to the sampling edge (before the 10^th^ or beyond the 90^th^ percentile of the sampling site elevations, depicted as orange rectangles in Fig. \ref{fig:fig3}), for example,  *Yersinella sp.* and *Gomphocerus sibiricus sibiricus*, or species having low maximum occupancy probability, e.g. *Omocestus haemorrhoidalis*.

@@

EHMOS always performed better than the t-test-based method. Differences between these two methods were particularly large under unbalanced sampling design, especially in the presence of generalist species. It is predictable that mean estimates are biased by unbalanced sampling designs, in particular in scenarios where a part of the gradient is preferentially sampled [@terbraak1986]. We simulated one example where the middle part of the gradient was preferentially sampled, but similar results are expected in other preferential sampling design. In such situations, both optima estimates could be biased towards the preferentially sampled part of the gradient and shift estimates could be biased towards zero. This explains the negative bias produced by t-test method in simulations involving unbalanced sampling design. It is also not surprising that bias due to unbalanced sampling design is larger for generalist species than for specialist species. Bias is more influenced by the distribution of sampling sites occupied by the species rather than by the entire distribution of sampling sites. Occupied sites of a specialist species with a narrow range could be uniformly distributed even if the entire distribution of sampling sites is uneven. In contrast, the larger the species range, the more the occupied site distribution on the gradient will match the entire sampling site distribution, leading to biased shift estimates.

@@

Generalized linear models have been extensively used in species distribution modelling [@elith2010], and in optimum shift modelling [e.g. @coudun2005; @lenoir2008]. Usually, multiple GLMs are conducted separately for each species and each survey that is to be compared [but see @lembrechts2017]. This  approach has the inconvenient of losing information and thus precision in estimation by splitting the data into multiple data sets. Estimates for data-poor species may be improved with a GLMM, by borrowing information from data-rich species [@ovaskainen2011]. Hence, one could assume that GLM would have, at best equal and probably worse performances than cGLMM.  We thus chose to compare EHMOS only with cGLMM to show that EHMOS has advantages against classical GLM-based approaches. Our Bayesian framework allows us to derive uncertainty measures associated with shift estimates. It would have been possible to compute the asymptotic variance of optima or shift estimates using a frequentist approach, e.g. the delta method [@serfling1980;@billingsley1986]. The Bayesian GLMM we developed performed as well as the EHMOS and better than the mean comparison method in scenarios where all species have optima in the middle of the sampling range. However, as expected, for edge species, the standard GLMM failed to estimate bell shape curves, fitting sigmoid curves instead and making derived optima and shifts unreliable [@citores2020; @terbraak1986]. In contrast, EHMOS, thanks to its ecological formulation and explicit modelling of optimum shifts, allowed estimates of optimum shifts for edge species. Hence, edge species could be kept if one is using EHMOS, while they should be removed in t-test or GLM-based analysis. Ecological formulation of EHMOS, especially explicit modelling of shift, also allows easier choice of prior in a Bayesian setting. cGLMM parameters should be transformed to be ecologically interpretable (see Appendix S1). Such a transformation may change an uninformative prior on the parameter scale into an informative prior on the ecological scale, making the choice of prior complicated [see @lemoine2019 for an example in logistic regression]. Hence, the standard GLMM appeared to be less flexible than EHMOS.

In addition to having more robust point estimate accuracy whatever the sampling design, ecological marginality and specialization than a t-test or cGLMM, EHMOS also seemed to produce more precise estimates. cGLMM was very sensitive to marginality in terms of precision, providing very wide and uninformative credible intervals for edge species. The interval coverage of the t-test was also very sensitive to marginality, sampling design and specialization. Having informative precision estimates is important when studying species shifts [@bates2015; @taheri2021]. Some authors use the overlap of confidence intervals as a surrogate for a test of significant difference [@lenoir2008; @urli2014], so low interval coverage could lead to misleading conclusions, and the t-test method should be avoided.

Here, we studied optimum shifts between two surveys conducted 30 years apart of 24 Orthoptera species along an elevational gradient in the Alps. All methods suggested that on average Orthoptera species shifted their optimum towards higher elevations, and that the magnitude of the shift varies between species. These results are consistent with previous findings on other insect taxa, which on average have shifted upslope, but at different paces depending on the species [@vitasse2021]. Even if the magnitude of shifts varied between methods, our estimates are in line with the estimated warming rate of 0.36 $^{\circ}$C.decade^-1^ in the Alps between 1970 and 2019, which roughly corresponds to a mean shift of about +62–71 m decade^–1^ (Vitasse et al. 2021), leading to an expected mean shift of around +186–213 m between the two surveys. Thus, the average shift estimated by EHMOS (`r appli_mean_shift[appli_mean_shift$model == "ehmos", ]$m.shift %>% round(0)` m) is closer to what is expected than the t-test, suggesting it may provide a more reliable estimate of optimum shift. Yet this observation has to be taken carefully, as the warming rate could vary between regions, and species might respond at a slower rate than expected [@vitasse2021]. However, as the case study is close to the simulated scenario including the presence of both generalist and edge species sampled with an unbalanced sampling design, we could suspect an underestimation of optimum shifts by the t-test. It was also not surprising to obtain unreliable credible intervals for most of the edge species with the GLMM, as in the simulations.

Notwithstanding its demonstrated comparative advantages, EHMOS could have some limitations. It might produce optimum estimates outside the sampling range and thus, depending on the sampling design, outside the environmental range. This property could be an advantage in the case of niche truncation. However, optimum estimates should be limited to the environmental range: for example, by imposing an appropriate prior. One potential other limitation, also valid for cGLMM, is the assumption that species response curves are symmetric. While this assumption could stand for numerous species [@rydgren2003], some species could have asymmetrical relationships with environment gradients [@oksanen2002]. We found through simulations that the three methods could be biased in presence of asymmetrical species response curves (see Appendix S4). More precisely, the three methods provided biased optimum shift estimates when species response curves changed of shapes between the sampling occasions. For instance, the three methods produced positively biased optimum shift estimates for species having a symmetrical response curve in the first sampling occasion and a right-skewed species response curve in the second sampling occasion. In contrast, optimum shift should be accurately estimated by the three methods for species having the same degree of skewness in both sampling occasions. Hence, shape of species response curves should be inspected carefully prior to analysis. In case where response curves are asymmetric, some alternative modelling methods have been proposed, such as generalized additive models [GAM, e.g. @heegaard2002] or Huisman-Olff-Fresco approach [HOF, e.g. @oksanen2002], and applied in optimum shift modelling [@maggini2011; @urli2014]. However, previous simulation studies have shown that both methods could fail to retrieve the shape of species response [@jansen2013; @michaelis2017]. Both methods appeared to be sensitive to gradient truncation and sampling intensity [@jansen2013; @citores2020], limiting their use in unbalanced sampling design and for edge species. Further research is needed to find more robust methods to deal with asymmetrical response curves. Finally, we present the EHMOS within a hierarchical model that did not incorporate observation process. Imperfect detection could bias occupancy estimates, and this should be addressed in modelling whenever it is possible. While our example does not allow the use of multi-species occupancy models, hierarchical linear models estimating both species occupancy and detection [@dorazio2005], due to the absence of sampling replication in historical data, it is straightforward to extend our framework in a MSOM by adding the observation process in the hierarchy of the model.
